---
title: "Application_example"
author: "Nicola Dalla Valle"
date: "2025-09-30"
output: html_document
---

```{r warnings=FALSE}
library(igraph)
library(dplyr)
library(Spectra)
library(tidyverse)
library(Spectra)
library(xcms)
library(MsExperiment)
library(MsCoreUtils)
library(purrr)
library(MASS) 
library(dbscan)
library(qgam)
```

#MS/MS extraction and dataset processing

loading files which include DDA experiments

```{r}
pth <- file.path( "alternaria_data/pos/data/mzML")
fls <- list.files(pth, pattern = "mzML", full.names = TRUE)
fls <- fls[grep("DDA", fls)] 
```

peak piking 
 
```{r}
mse <- readMsExperiment(spectraFiles = fls)

#filter wrong spectra
idx <- which(round(spectra(mse)$precursorMz) != round(spectra(mse)$isolationWindowTargetMz) & spectra(mse)$msLevel==2)
filterByIndex <- function(x, index) { x[index] }
filterSpectra(mse, filterByIndex, -idx)

#peak piking
cwp<- CentWaveParam(ppm = 8,
                    peakwidth = c(3, 12),
                     prefilter = c(6, 5000000),
                     noise = 100000)
data <- findChromPeaks(mse, param = cwp) 

#indicating the type of sample
sampleData(data)$sample_type <- "study"
sampleData(data)$sample_type[grep("QC",sampleData(data)$spectraOrigin)] <- "QC"
sampleData(data)$sample_type[grep("MEDW",sampleData(data)$spectraOrigin)] <- "MED"
sampleData(data)$sample_type[grep("blank",sampleData(data)$spectraOrigin)] <- "blank"
```

peak filtering based on signal to noise ratio
 
```{r}
pks <- as.data.frame(chromPeaks(data))
keep_peaks <- rownames(pks)[pks$sn > 25]  #filtre for s/n ratio
data <- filterChromPeaks(data, peakid = keep_peaks)
```
 
features creation
 
```{r}
#' Peak refinement
mnp <- MergeNeighboringPeaksParam(expandRt = 1, expandMz = 0.001,
                                  minProp = 3/4)
data <- refineChromPeaks(data, param = mnp)

#' Alignment
pdp1 <- PeakDensityParam(sampleGroups = sampleData(data)$sample_type, bw = 2,
                         minFraction = 1, binSize = 0.01)
data <- groupChromPeaks(data, param = pdp1)

data <- adjustRtime(
  data, 
  param = PeakGroupsParam(minFraction = 1,
                          subset = which(sampleData(data)$sample_type == "QC"),
                          subsetAdjust = "average", 
                          span = 0.2))

#' Correspondence analysis
pdp2 <- PeakDensityParam(sampleGroups =sampleData(data)$sample_type, bw = 2,
                         minFraction = 0.7, binSize = 0.01) 
data <- groupChromPeaks(data, param = pdp2)

#' Gap-filling
data <- fillChromPeaks(data, param = ChromPeakAreaParam())

```

feature filtering to consider "real" feature from biological sample

```{r}
xdata<- data
#rt limitation
data <- filterRt(data, rt=c(30,660))

#filtering for missin values in Qcs
f <- sampleData(data)$sample_type
f[f != "QC"] <- NA
missing_filter <- PercentMissingFilter(threshold = 20, f = f)
data <- filterFeatures(object = data, filter = missing_filter)

#filtering features with similar intensity to blank 
blank_flags <- BlankFlag(threshold = 2,
                    qcIndex = sampleData(data)$sample_type == "QC",
                    blankIndex = sampleData(data)$sample_type == "blank")

data <- filterFeatures(data, blank_flags)

#elminate feature with similar  variability  to the QCs (likely noise)
dratio_filter <- DratioFilter(
    threshold = 0.5,
    qcIndex = sampleData(data)$sample_type == "QC",
    studyIndex = sampleData(data)$sample_type == "study",
    na.rm = TRUE)

# Apply the filter to object
data <- filterFeatures(object = data,
                                  filter = dratio_filter)

ftvalues <-data.frame(featureValues(data, value = "into", method="sum"))
ftdetec <- data.frame(featureDefinitions(data))
ftdetec$name <- rownames(ftdetec)
ftvalues <- ftvalues[rownames(ftvalues) %in% rownames(ftdetec),]
ftdetec$meanval <- colMeans(t(ftvalues), na.rm = TRUE)

```

searching for msms related to each features
 
```{r}
ms2_spectra <- featureSpectra(data, expandRt = 0, return.type = "Spectra",
                              method ="largest_bpi",
                              ppm = 5, 
                              featureColumns = c("mzmed", "rtmed", "rtmin", "rtmax"))

ms2_spectra$id <- paste0(basename(ms2_spectra$dataOrigin), ms2_spectra$scanIndex)

#features with at leat one fragmentation spectra
features <- unique(spectraData(ms2_spectra)$feature_id)

```

selecting one ms2 for each feature, selecting the one with highest base peak intensity

```{r}

sp <- split(spectraData(ms2_spectra), spectraData(ms2_spectra)$feature_id)

cp <- vapply(sp, function(x){
    x$chrom_peak_id[which.max(x$basePeakIntensity)]
}, character(1))

ms2_spectra <- ms2_spectra[which(spectraData(ms2_spectra)$chrom_peak_id %in% cp)]
```

at this point I have one fragmentation spectra for each selected and filtered feature, relative only to real biological sample features.

#MS/MS similarity 

making matrix calculation applying GNPS method(considering both m/z and neutral loss matches).
prior Cosine calaualtion, msms are cleaned and precursor ion, if present, is removed.

```{r}
#Noise cleaning
#function to clean msms
cleaning_ref<- function(ms2){
  ms2_tb <- as.data.frame(
    spectraData(ms2) 
  ) %>%
  mutate(ms2_raw = map2(mz(ms2), intensity(ms2), ~{
    if (length(.x) == 0 || length(.y) == 0)
      return(data.frame(mz = numeric(0), intensity = numeric(0)))
    data.frame(mz = as.numeric(.x), intensity = as.numeric(.y))
  })) %>%
    mutate(ms2_RLM = purrr::map(ms2_raw, function(t){
         if (is.null(t) || nrow(t) == 0 || !"intensity" %in% names(t))
        return(tibble(mz = numeric(0), intensity = numeric(0)))
      t <- t %>%
        filter(intensity>0) %>%
        arrange(intensity) %>%
        mutate(order = seq(nrow(.)))
      p75 <- t %>%
        pull(intensity) %>% 
        quantile(0.75)
      u <- t %>%
        filter(intensity < p75)
      if(nrow(u)>3){
        mdl <- rlm(data = u, log10(intensity) ~ order)
        t_filt<- t %>% 
          mutate(pred = predict(mdl, t)) %>%
          mutate(res = abs(pred - log10(intensity))) %>%
          filter(res > sd(residuals(mdl))*3) %>%
          arrange(mz)
        if (nrow(t_filt) > 0) {
          t<- t_filt
        } else {
          t<- t %>% arrange(mz)
        }
      } else {
        return(t %>% arrange(mz))  # Not enough points to model
      }
    }))
  ms2_cl <- ms2_tb %>%
    mutate(
      mz = purrr::map(ms2_RLM, ~ .x$mz),
      intensity = purrr::map(ms2_RLM, ~ .x$intensity)
    ) %>%
    filter(map_lgl(mz, ~ length(.x) > 0)) %>%
    as.data.frame() %>%
    Spectra()
  return(ms2_cl)
}

spspos <- cleaning_ref(ms2_spectra)

ftpos <- ftdetec[which(ftdetec$name %in% spspos$feature_id),]
fts <-ftpos$name
sps <- spspos[which(spspos$feature_id %in%fts )]

#Remove precursor peaks 
spc <- filterPrecursorPeaks(sps, ppm = 50, mz = ">=") %>%
   scalePeaks()
spc <- spc[lengths(spc)>0]

#similarity claualtion
matrix <- compareSpectra(spc, MAPFUN = joinPeaksGnps, tolerance = 0.01, FUN = MsCoreUtils::gnps)
colnames(matrix)<- spc$feature_id
rownames(matrix)<- spc$feature_id
```

#MN NLs and FIs informativity 

function for spectral features informativity ranking

```{r}
#extracting "MN_info" function
source("MN_info.R")
```

application of the function to NLs and FIs 
Then, I consider both FIs and NLs to evaluate overall influence on MN
 
```{r}
#neutral losses
nls <- MN_info(method="NLs",matrix, spc, perm=100, thr= 0.7, min_presence = 10)

#fragment ions
mzs <- MN_info(method="FIs",matrix, spc, perm=100, thr= 0.7, min_presence = 10)

#merging the results 
nls <- nls %>%
  dplyr::select(value, number, z_score, excess) %>%
  mutate(origin="NL") %>%
  mutate(log_number= log10(number))

mzs <- mzs %>%
  dplyr::select(value, number, z_score, excess) %>%
  mutate(origin="mz")%>%
  mutate(log_number= log10(number))

alldf <- rbind(nls, mzs) %>%
  arrange(desc(excess))
```

visualizing MN with specific FIs of NLs highlited

```{r}
library(visNetwork)

#extracion of NLs for all msms
neutral_loss <- function(x, precursorMz, ...) {
    # Keep only peaks with at least 5% of the base peak intensity
    base_intensity <- max(x[, "intensity"], na.rm = TRUE)
    x <- x[x[, "intensity"] >= 0.00 * base_intensity, , drop = FALSE]
    
    # Compute neutral loss
    x[, "mz"] <- precursorMz - x[, "mz"]
    
    # Sort by neutral loss m/z
    x[order(x[, "mz"]), , drop = FALSE]
  }
  nl <- addProcessing(msms, neutral_loss,
                      spectraVariables = "precursorMz")
  msms$nl<- mz(nl)
  
#igrah object creation
  
mat_thr <- matrix
mat_thr[mat_thr < 0.7] <- 0    

g <- graph_from_adjacency_matrix(
  mat_thr,
  mode = "undirected",
  weighted = TRUE, 
  diag = FALSE
)
g <- delete_vertices(g, V(g)[degree(g) == 0])

#nodes creation
nodes <- data.frame(
  id=as.character(unique(V(g)$name))
)
idx <- match(nodes$id, msms$feature_id)

nodes$mz<- precursorMz(msms)[idx]
nodes$rt <- rtime(msms)[idx]/60 
nodes$frag <- mz(msms)[idx]
nodes$nl <- msms$nl[idx]

set.seed(123)  
layout_fr <- layout_with_fr(g)

nodes$x <- layout_fr[,1]*100
nodes$y <- layout_fr[,2]*100

#MATCHES
tol <- 0.005

#for neutral losses
nl_targets <- c(123.456)
nodes$has_all_nl <- map_lgl(nodes$nl, function(vals) {
  vals <- as.numeric(as.character(vals))
  all(sapply(nl_targets, function(nli) any(abs(vals - nli) <= tol, na.rm = TRUE)))
})

#for mz vlaues
mz_targets <- c(123.456)
nodes$has_all_mz <- map_lgl(nodes$frag, function(vals) {
  vals <- as.numeric(as.character(vals))
  all(sapply(mz_targets, function(nli) any(abs(vals - nli) <= tol, na.rm = TRUE)))
})

#colours according to mz matches
edges <- igraph::as_data_frame(g, what = "edges")
nodes$color.background <- case_when(
  nodes$has_all_nl ~ "navy",
  nodes$has_all_mz ~ "indianred",
  TRUE             ~ "grey"
)

#MN plotting
network_plot <- visNetwork(nodes %>%
                             dplyr::select(-c(nl, frag)), edges) %>%
 visNodes(
    x = ~x,
    y = ~y,
    fixed = TRUE,
    label = NULL,
    shape = "dot",
    borderWidth = 0,
    color = list(border = "", highlight = "red")
  ) %>%
  visEdges(
    color = list(color = "lightgray", highlight = "blue"),
    width = 2
  ) %>%
  visOptions(
    highlightNearest = list(enabled = TRUE, degree = 1, hover = TRUE),
    nodesIdSelection = FALSE
  ) %>%
  visPhysics(enabled = FALSE)


network_plot
```


